---
title: "Tidytuesday: Global Health Spending"
author: "Thomas Chula"
date: "2026-04-24"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.align = "center",
  dpi = 300
)

pacman::p_load(tidytuesdayR, tidyverse, janitor, skimr, tidylog, kableExtra, scales, plotly)

```

```{r attach data}
tuesdata <- tuesdata <- tidytuesdayR::tt_load('2026-04-21')
financing_schemes <- tuesdata$financing_schemes
health_spending <- tuesdata$health_spending
spending_purpose <- tuesdata$spending_purpose
```
# Introduction
The TidyTueday Data for week 16 (released on `2026-04-21`) is about Global Health Spending. It comprises of 195 countries and territories since 2000. 
The data tracks how much countries spend on health care, how money comes in and how it is channelled through financing schemes, and how it is spent on. The data is organized into three datasets.

- Health spending: Aggregate health spending and its breakdown by funding source (domestic government, domestic private, and external aid).

- Financial Schemes: Health spending by financial scheme such as government schemes, voluntary insurance, and out of pocket payments

- Spending purpise: Health spending by healthcare function such as curative care, preventive care and medical goods.

# Executive Summary

# Exploratory Data Analysis
## Financial Schemes
The `Financial Schemes`dataset contains data of `r nrow(distinct(financing_schemes, country_name))` countries on `r nrow(distinct(financing_schemes, financing_scheme))` types of financing schemes. 
The different types of schemes are displayed below

```{r eda financial_schemes}
countries <- financing_schemes %>% 
  count(financing_scheme) %>% 
  select(-n, 'Financing Scheme' = 'financing_scheme')

kable(countries)

```
### General Breakdown of Financing Scheme
According to the chart and table, it is visible that `Household Out of Pocket payment (OOPS)` scheme has the highest contribution, whereas `Rest of the world financing schemes (non-resident)` Scheme has contributed the least.
```{r oops}
financing_value <- financing_schemes %>% 
  filter(!str_detect(indicator_code, 'che')) %>% 
  # filter(str_detect(financing_scheme, 'Government|pocket|Rest|Unspecified|Voluntary')) %>% 
  # filter(!str_detect(unit, '%')) %>% 
  group_by(financing_scheme) %>% 
  summarise(
    avg_spending = dollar(mean(value)),
    median_spending = dollar(median(value)),
    minimum_value = dollar(min(value)),
    maximum_value = dollar(max(value))
  ) %>% 
  rename()

kable(financing_value, caption = 'Breakdown of financial values by financial scheme') %>% 
  kable_classic()



financing_value %>% 
  select(financing_scheme, avg_spending) %>% 
  mutate(`financing scheme` = case_when(
    str_detect(financing_scheme, 'Government') ~ 'Government Scheme',
    str_detect(financing_scheme, 'Household') ~ 'OOPS',
    str_detect(financing_scheme, 'Rest') ~ 'Non-Resident',
    str_detect(financing_scheme, 'n.e.c') ~ 'Unspecified',
    str_detect(financing_scheme, 'Voluntary') ~ 'Volunrary HC Payments'
  )) %>% 
  ggplot(aes(`financing scheme`, avg_spending)) + 
  geom_col(fill = "steelblue") +
  geom_text(aes(label = avg_spending),
            vjust = -0.5,
            size = 3) +
  labs(title = "Average Spending by Financing Scheme") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, color = "gray50")
  )
```
## Trend of Spending by scheme
The trend of spending by scheme is displayed below

```{r}
spending <- financing_schemes %>% 
  filter(!str_detect(indicator_code, 'che')) %>% 
  mutate(value_billions = value / 1e9) %>% 
  slice_head(n = 50)
  

ggplot(spending, aes(year, dollar(value_billions))) + geom_col() + labs(title = 'Financing Scheme Trend', y = 'Value in Billions', x = 'year')


```
